Learning from Data Through Models

### 2023-07-11 paper/paper.Rmd
writing predictions

---
title: "Learning from Data Through Models"
date: '`r format(Sys.time(), "%B %d, %Y")`'
author: "Alberto Bisin, Guillaume Frechette and Jimena Galindo"
abstract: "TBW"
output: 
    pdf_document:
        citation_package: natbib
fontsize: 12pt
geometry:
- margin=2.5cm
hyperref:
- linestretch=2.0
documentclass: article
---

@@

A model of the world is said to be *simple* if it does not include all the relevant variables, 
*i.e.* if $M\subset \mathcal{R}$. And it is complex if it includes all relevant variables, $M=\mathcal{R}$.

@@

## Polarization
To study polarization, we take the definition of CITATION: two DMs are said to be 
polarized if, when observing the same realization of observables, they predict the state to be different. That is, 
they both observe the same draw of $x=(x_1, ..., x_N)$ but one DM predicts the state to be **B** and the other 
predicts the state to be **R**.

Within out framework---as in CITATION---polarization may persist even when DMs have 
access to infinite data because they might be using different models of the world in order to make their predictions. 
Nevertheless using different models does not necessarily imply that beliefs will always be polarized; it must also be 
the case that the realization of observables falls in a *disagreement zone*. By disagreement zone, we simply 
mean that if the two models that are being used for prediction are $M_1$ and $M_2$, the realization of the 
observables $x=(x_1, ..., x_N)$ must be such that $M_1$ makes a different prediction from $M_2$. The following 
2-dimensional example illustrates why the use of different models is necessary but not sufficient for polarization 
to arise.

@@

For any pair of agents who observe the same realization of teh observables and use different models, it is possible
to determine whether they will be polarized or not by looking at the predictions made by their models.

@@

Subjects for the experimet were recruited from the CESS lab subject pool at NYU. 
We recruited `r data[, uniqueN(participant.code)]` undergraduate students who participated in this experiment in person at 
the CESS lab. The experiment was coded using o-tree CITATION and it consisted of two parts: Part 1 was meant to expose the 
subjects to the data generating process
and allow them to learn how to predict; part 2 was meant to elicit the models of the world that subjects might have
developed in part 1.

@@

Part 2 started immediatly after part 1. In this part each subject was randomly allowed to use only a certain number of variables
in order to make their prediction. Instead of showing them the realizations of all 5 variables, they were told that they can 
disclose up to $K$ variables, where $K$ was drawn at random from the set $\{1, 2, ..., 5\}$ at every round and it was drawn 
independently across subjects. We take the variables that they chose to reveal as the model that each subject using for predicting 
the state and use that model to detrmine whether the theory would predict the pair to be polarized or not. In addition, we use 
the random assignment of $K$ to investigate whether the subjects are using siumple or complex models. If allowing subjects to 
disclose an additional variable improves their prediction accuracy, then it must be that they are in fact using the additional
information for prediciton. If, on the other hand, they do not perform better when they have access to more information, it must be
that although they are disclosing an additional variable, they do not correctly incorporate the infomation into their prediciton 
rule. Thus, if we observe that the performance of our subjects does not improve when they are allowed to disclose $K+1$ variables, 
relative to their performance when they were allowed to disclose only $K$ variables, their model of the world must be of size $K$
or lower.

@@

# References

@@

Inline code example: `r data[, uniqueN(participant.code)]`

>>>>

---
title: "Learning from Data Through Models"
date: '`r format(Sys.time(), "%B %d, %Y")`'
author: "Alberto Bisin, Guillaume Frechette and Jimena Galindo"
abstract: "TBW"
output: 
    pdf_document:
        citation_package: natbib
bibliography: bibliography.bib
fontsize: 12pt
geometry:
- margin=2.5cm
hyperref:
- linestretch=2.0
documentclass: article
---

@@

A model of the world is said to be *simple* if it does not include all the relevant variables, 
*i.e.* if $M\subset \mathcal{R}$. And it is complex if it includes all relevant variables, $M=\mathcal{R}$. Thus, whenever 
agents are using simple models, they are ignoring some of the relevant variables even if the information is available to them.
In contrast, whenever agents are using complex models, they are using all the relevant information that is available to them, 
so restricting the number of variables that is available to them will impair their ability to predict correctly. In the design
of the experiment we will exploit these differences in order to determine whether subjects are using simple or complex models.

**Prediction 1:** If subjects are using simple model with $K$ variables, then allowing them to disclose $K+1$ variables 
will not increase the frequency with which they make correct predictions.

**Prediction 1B:** If subjects are using complex models, then allowing them to disclose an additional variable will always 
increase the frequency with which they make correct predictions.

@@

## Polarization
To study polarization, we take the definition of @Haghtalab2021: two DMs are said to be 
polarized if, when observing the same realization of observables, they predict the state to be different. That is, 
they both observe the same draw of $x=(x_1, ..., x_N)$ but one DM predicts the state to be **B** and the other 
predicts the state to be **R**.

Within out framework---as in @Haghtalab2021---polarization may persist even when DMs have 
access to infinite data because they might be using different models of the world in order to make their predictions. 
Nevertheless using different models does not necessarily imply that beliefs will always be polarized; it must also be 
the case that the realization of observables falls in a *disagreement zone*. By disagreement zone, we simply 
mean that if the two models that are being used for prediction are $M_1$ and $M_2$, the realization of the 
observables $x=(x_1, ..., x_N)$ must be such that $M_1$ makes a different prediction from $M_2$. The following 
2-dimensional example illustrates why the use of different models is necessary but not sufficient for polarization 
to arise.

@@

For any pair of agents who observe the same realization of teh observables and use different models, it is possible
to determine whether they will be polarized or not by looking at the predictions made by their models. Notice that the 
polarizing observations---those that fall into a polarization zone---are different for each pair of models. Also notice that 
our definition of the polarization zone assumes that the DMs are using optimal rules. This is a very strong assumption 
and will most likely not be the case that subjects in the experiment learn perfectly. However, the notion of optimal rules
allows us to have a clean definition of polarization and to make predictions about when polarization will arise.

**Prediction 2 :** Polarization will arise more often when the theory predicts so. That is, when two agents who have 
different models of the world, are facing a realization of observables that falls into the polarization zone corresponding 
to their models.

@@

Subjects for the experimet were recruited from the CESS lab subject pool at NYU. 
We recruited `r data[, uniqueN(participant.code)]` undergraduate students who participated in this experiment in person at 
the CESS lab. The experiment was coded using o-tree @otree and it consisted of two parts: Part 1 was meant to expose the 
subjects to the data generating process
and allow them to learn how to predict; part 2 was meant to elicit the models of the world that subjects might have
developed in part 1.

@@

Part 2 started immediatly after part 1. In this part each subject was randomly allowed to use only a certain number of variables
in order to make their prediction. Instead of showing them the realizations of all 5 variables, they were told that they can 
disclose up to $K$ variables, where $K$ was drawn at random from the set $\{1, 2, ..., 5\}$ at every round and it was drawn 
independently across subjects. We take the variables that they chose to reveal as the model that each subject using for predicting 
the state and use that model to detrmine whether the theory would predict the pair to be polarized or not. In addition, we use 
the random assignment of $K$ to investigate whether the subjects are using siumple or complex models. If allowing subjects to 
disclose an additional variable improves their prediction accuracy, then it must be that they are in fact using the additional
information for prediciton. If, on the other hand, they do not perform better when they have access to more information, it must be
that although they are disclosing an additional variable, they do not correctly incorporate the infomation into their prediciton 
rule. Thus, if we observe that the performance of our subjects does not improve when they are allowed to disclose $K+1$ variables, 
relative to their performance when they were allowed to disclose only $K$ variables, their model of the world must be of size $K$
or lower. This feature design allows us to test prediciton 1 and 1B and determine the number of variables that subjects are using.

@@

@@
